---
title: "hw1"
author: "Jack"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
setwd("C:/Users/jackw/Documents/R")
```
##########
# QUESTION
# NUMBER 1
##########

# reading table (q1)
```{r}
library(readr)
indicators <- read_table("datasets/indicators.txt")
View(indicators)
```

# running the linear model function on indictors.csv (q1)
```{r}
library(ISwR)
fit <- lm(PriceChange ~ LoanPaymentsOverdue, indicators)
summary(fit)
```
############
## Q2.2 (A) 
############

# finding the inital confidence interval for the lm (q1.1)
```{r}
confint(fit, 'LoanPaymentsOverdue', level=0.95)
paste("There is evidence of a significant, negative linear association because it's highly proabable that independent variable LoanPaymentsOverdue is negative.")
```


############
## Q2.2 (B) 
############

# using the lm to estimate E(Y|X=4) and find a predicted 0.95 confidence interval (q1.2)
```{r}
xvalue <- 4
# predict is best used when setting independent variable values.
predict(fit, data.frame(LoanPaymentsOverdue = xvalue), interval = "confidence")
paste("0% is NOT a feasible value for E(Y|X=4) because the upperbound of the confidence interval is below the zero value.")
```
##########
# QUESTION
# NUMBER 2
##########

# reading table (q2)
```{r}
library(readr)
invoices <- read_table("datasets/invoices.txt")
View(invoices)
```
# running the linear model function on invoices.csv (q1)
```{r}
library(ISwR)
fitq2 <- lm(Time ~ Invoices, invoices)
summary(fitq2)
```
############
## Q2.3 (A) 
############

# manually computing confidence interval (q2.1)
# confint() only works for independent variables and not for model parameters i.e. b0.
```{r}
confint(fitq2, '(Intercept)', level=0.95)
```

############
## Q2.3 (B) 
############

```{r}
# t-test
# h0 -> B1 = 0.01
# h0 -> B1 != 0.01
invoice_benchmark <- 0.01
invoice_estimated_value <- 0.0112916
invoice_standard_error <- 0.0008184
# how many standard errors the estimated value is from the hypothesized value
invoice_t_value <- (invoice_estimated_value - invoice_benchmark) / invoice_standard_error
paste("t_value:", invoice_t_value)

# degrees of freedom = sample size - parameters being estimated
degrees_of_freedom <- 28

# probability of observing a t-value as extreme as the one calculated, assuming the h0=TRUE.
invoice_p_value <- 2 * pt(abs(invoice_t_value), degrees_of_freedom, lower.tail = FALSE)
paste("p_value:", invoice_p_value)
paste("We fail to reject the null hypothesis because the p-value was found to be ~0.12, greater than the standard significance level of 0.05. So, we cannot assume that the average processing time is signifcantly different than the benchmark of 0.01")
```

############
## Q2.3 (C) 
############

# point estimate (q2.3)
```{r}
intercept_coef <- coef(fitq2)[1]
invoices_coef <- coef(fitq2)[2]
invoice_num <- 130
# Y = B0 + B1 * Xi
point_estimate <- intercept_coef + invoices_coef * invoice_num
paste("point estimate:", point_estimate)
```
# 95% prediction interval (q2.3)
```{r}
predict(fitq2, newdata = data.frame(Invoices = c(130)), interval = "prediction", level = 0.95)
``` 