---
title: "Project"
author: "Jack"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup}
knitr::opts_chunk$set(warning = FALSE)
```

Facebook Social Media Metrics: Use facebook_updated.csv that contains the same data set
but includes an additional column obs_type, indicating whether each observation should be
considered for training or validation. Your goal is to construct a model using training data
(obs_type == “Training”) and predict the ‘Lifetime Post Consumers’ for the validation
observations (obs_type == “Validation”). To do this, you'll use the training observations to
train your linear model. Once you have finalized a model, use it to make predictions for the
validation data set. Consider the seven input features category, page total likes, type,
month, hour, weekday, paid as predictors. Leverage your findings from lab report #3 to
inform which combination of these predictors might work best. You should aim to build a model
that can make predictions as accurately as possible.

```{r}
setwd("C:/Users/jackw/Documents/R/Project2")
# install.packages("car")
# install.packages("ggplot")
library(car)
library(ggplot2)
# library(GGally)
```


```{r}
library(readr)
facebook <- read_delim("facebook_updated.csv", 
    delim = ",", escape_double = FALSE, trim_ws = TRUE)
View(facebook)
```

```{r}
# type factor 
facebook$`Type` <- factor(facebook$`Type`)
str(facebook$`Type`)

# month factor
month_names <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun" , "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")
facebook$`Post Month` <- factor(facebook$`Post Month`, levels = 1:12, labels = month_names)
str(facebook$`Post Month`)

facebook <- na.omit(facebook)
```


```{r}
# normalizing likes
facebook$like[is.na(facebook$like)] <- 0
facebook$log_like <- NA
facebook$log_like <- log(facebook$`like`)
facebook$like[facebook$like=='-Inf'] <- 0

# normalizing `Lifetime Post Total Reach`
facebook$lt_post_tot_reach <- NA
facebook$lt_post_tot_reach <- log(facebook$`Lifetime Post Total Reach`)

facebook <- na.omit(facebook)
facebook <- facebook[is.finite(facebook$`Lifetime Post Total Reach`) & is.finite(facebook$log_like) & is.finite(facebook$`Page total likes`), ]

lm_model <- lm(lt_post_tot_reach ~ `Type` + `Page total likes` + I(log_like^1.5), data = facebook)

# plot(facebook$`Page total likes`, facebook$`Lifetime Post Total Reach`)
plot(facebook$log_like^2, facebook$`Lifetime Post Total Reach`)
plot(facebook$`like`, facebook$`Lifetime Post Total Reach`)
# 
# summary(facebook$`Page total likes`)
summary(lm_model)
plot(lm_model)
vif(lm_model)
avPlots(lm_model)
```

```{r}
# normalizing likes
facebook$like[is.na(facebook$like)] <- 0
facebook$log_like <- NA
facebook$log_like <- log(facebook$`like`)
facebook$like[facebook$like=='-Inf'] <- 0

# normalizing `Lifetime People who have liked your Page and engaged with your post`
facebook$lt_ppl_lkd_egnged <- NA
facebook$lt_ppl_lkd_egnged <- log(facebook$`Lifetime People who have liked your Page and engaged with your post`)

facebook <- na.omit(facebook)
facebook <- facebook[is.finite(facebook$`Lifetime Post Total Reach`) & is.finite(facebook$log_like) & is.finite(facebook$`Page total likes`), ]

lm_model <- lm(lt_ppl_lkd_egnged ~ `Type` + `Page total likes` + log(`Lifetime Post Total Impressions`) + I(log_like^2), data = facebook)

summary(lm_model)
plot(lm_model)
vif(lm_model)
avPlots(lm_model)
```


```{r}
# str(facebook)

facebook$Paid <- as.factor(facebook$Paid)

# post reach
facebook$log_ltpostreach <- NA
facebook$log_ltpostreach <- log(facebook$`Lifetime Post reach by people who like your Page`)
plot(density(facebook$log_ltpostreach))

# likes
facebook$like[is.na(facebook$like)] <- 0
facebook$log_like <- NA
facebook$log_like <- log(facebook$`like`)
plot(density(facebook$log_like))

# workday
facebook$hours_factor <- ifelse(facebook$`Post Hour` >= 9 & facebook$`Post Hour` < 17, "Work", "Non-Work")
facebook$hours_factor <- factor(facebook$hours_factor, levels = c("Work", "Non-Work"))

# total interactions
facebook$log_tot_interact <- NA
facebook$log_tot_interact <- log(facebook$`Total Interactions`)
plot(density(facebook$log_tot_interact))

facebook$like[is.na(facebook$share)] <- 0
facebook$log_share <- NA
facebook$log_share <- log(facebook$share)
plot(density(facebook$log_share))

facebook <- na.omit(facebook)
facebook <- facebook[is.finite(facebook$share) & is.finite(facebook$log_ltpostreach) & is.finite(facebook$log_like), ]

lm_model3 <- lm(share ~ log_like + log_tot_interact + hours_factor, data = facebook)

residuals_lm <- residuals(lm_model3)
threshold <- 50 
high_residuals <- which(abs(residuals_lm) > threshold)
facebook_cleaned <- subset(facebook, !row.names(facebook) %in% high_residuals)

lm_model_cleaned <- lm(log_share ~ log_tot_interact + log_like + hours_factor + I(log_like^2) + I(log_tot_interact^2), data = facebook_cleaned)

summary(lm_model_cleaned)
plot(lm_model_cleaned)
```