---
title: 'BS755 / MA575 Lab 3: Data Visualization'
date: "Friday, September 22, 2023"
output:
  html_document:
    toc: yes
  pdf_document:
    toc: yes
---

# Lab 3: Data Visualization

In this lab, we will explore data visualization using an air quality data set.

## Setup

# Setting up global options to supress printing any warnings. Install `rmarkdown` and `knitr` packages.

```{r setup}
# Global options to suppress warnings in all code chunks
knitr::opts_chunk$set(warning = FALSE)

# Install rmarkdown and knitr packages if they are not already installed.
#install.packages("rmarkdown")
#install.packages("knitr")
```

### Set Working Directory

# First, set the working directory to the location where your data is stored on your computer. Uncomment and modify the `setwd` line as needed.

```{r}
setwd("C:/Users/jackw/Documents/R/Lab3")
```

### Install and Load Packages

Install and load the necessary R packages if they are not already installed. Uncomment the `install.packages` lines if needed.

```{r}
#install.packages("car")
#install.packages("ggplot2")
#install.packages("GGally")
library(car)
library(ggplot2)
library(GGally)
```

## Data Import and Exploration

### Read Data

Let's start by reading the data from a CSV file using a semicolon as the delimiter.

```{r}
# Read data from a CSV file using semicolon as delimiter
library(readr)
aq_data <- read_delim("AirQualityData(3).csv", delim = ";", escape_double = FALSE, trim_ws = TRUE)
View(aq_data)
```

### Check Missing Values

We'll check for missing values in the "CO.GT." column.

```{r}
# Plot Ground CO
plot(aq_data$`CO(GT)`)

# Summary of each column
summary(aq_data)

# In this data set it looks like missing values are stored as -200
# Check which values in Ground CO are -200
# which(aq_data$CO.GT. == -200)
# The output gives indices/positions in the vector with values -200

# Check how many values in Ground CO are -200
length(which(aq_data$CO.GT. == -200))

# Exercise: Check other columns in the data set for missing values 
```

### Data Cleaning

Replace missing data (-200) with NA.

```{r}
# Replace missing data (-200) with NA
aq_data[aq_data == -200] <- NA
# Note that this replaces -200 to NA across all columns in aq_data
```

### Data Type Conversion

Check for column types for the columns in the data set.

```{r}
# Check column types for all columns in the data set
str(aq_data)
# Check column types for specific columsn
# typeof(aq_data$Temperature)
```

Force specific columns to be numeric if they are not already.

```{r}
# Force specific columns to be numeric if not numeric/integer already.
# Note that change in variable names of PT08.S1.CO. and CO.GT.
Temperature <- as.numeric(as.character(aq_data$Temperature))
RelativeHumidity <- as.numeric(as.character(aq_data$RelativeHumidity))
AbsoluteHumidity <- as.numeric(as.character(aq_data$AbsoluteHumidity))
SensorCO <- as.numeric(as.character(aq_data$`PT08.S2(NMHC)`))
GroundCO <- as.numeric(as.character(aq_data$`CO(GT)`))
```

## Data Visualization

### Scatterplot of Ground CO

Let's create a scatterplot of Ground CO.

```{r}
# Scatterplot of Ground CO
plot(GroundCO)
```

### Scatterplot of Ground CO vs. Sensor CO

Now, we'll create a scatterplot of Ground CO vs. Sensor CO.

```{r}
# Plot scatterplot of GroundCO vs SensorCO
plot(SensorCO, GroundCO)

# Plot scatterplot of GroundCO vs SensorCO using the scatterplot function ...
# ... from the car package
scatterplot(SensorCO, GroundCO)

# Documentation for scatterplot function
# ?scatterplot

# Scatterplot of Ground CO vs. Sensor CO
plot(SensorCO, GroundCO,
     ylab="Ground Truth (CO)", xlab="Sensor CO",
     pch=19, cex=0.1)
```

### Scatterplot Matrix

We can also create a scatterplot matrix to visualize relationships among variables.

```{r}
# Scatterplot matrix of all the variables using the scatterplotMatrix ...
# ... function from the car package
scatterplotMatrix(~ GroundCO + SensorCO + Temperature + AbsoluteHumidity +
                    RelativeHumidity,
                  pch=19, cex=0.1)
```

### Scatterplot using ggplot2

Let's create a scatterplot using the `ggplot2` package.

```{r}
# Scatterplot using ggplot2
# Create a data frame with SensorCO and GroundCO
co_data <- data.frame(SensorCO, GroundCO)
# ggplot function for scatterplot
ggplot2(co_data, aes(x=SensorCO, y=GroundCO)) + 
  geom_point(size=0.1)
```

### Scatterplot Matrix using ggplot2

We can create a scatterplot matrix with correlation coefficients using `ggplot2`.

```{r}
# Create a data frame with GroundCO, SensorCO, Temperature, ...
# ... AbsoluteHumidity, RelativeHumidity
co_temp_data <- data.frame(GroundCO, SensorCO, Temperature, AbsoluteHumidity, RelativeHumidity)

# Scatterplot matrix with 2D contours
ggpairs(co_temp_data,
        lower=list(continuous=wrap("density", alpha=0.5), combo="box"),
        upper=list(continuous=wrap("points", alpha=0.3, size=0.1)))

# Scatterplot matrix with correlation coefficients
ggpairs(co_temp_data,
        upper=list(continuous=wrap("points", alpha=0.3, size=0.1)),
        lower=list(continuous=wrap('cor', size=4)))
```

This concludes our data visualization lab. Remember to save this Rmd script and use an R Markdown editor or RStudio to knit it into a PDF. Ensure you have [LaTeX](https://www.latex-project.org/) installed to compile this script directly into a PDF instead of HTML; you will need to change the option of `html_document` to `pdf_document` in the header.