---
title: 'BS755 / MA575 Lab 6: Multiple Linear Regression'
date: "Friday, October 13, 2023"
output:
  pdf_document:
    toc: yes
    toc_depth: '2'
  html_document:
    toc: yes
    toc_depth: 2
---
  
# Data Processing
  
  This Rmarkdown document presents the analysis of multiple linear regression on air quality data.

Set the working directory

```{r setup, include=FALSE}
# Set the working directory to access the data
# setwd('~/OneDrive - Boston University/Fall 2023/Teaching/Shariq BS755 MA575/Labs/Lab 06/')
```

Load the required packages

```{r load-packages, include=TRUE}
# Load required packages if not already installed
library(ggplot2)
library(dplyr)
library(GGally)
```

Reading and cleaning the data

```{r read-clean-data, include=TRUE}
# Read data from a CSV file using semicolon as delimiter
# Use '..' to move up one directory level to access the 'Lab 03' directory ...
# ... and read the 'AirQualityData.csv' file.
aq_data <- read.csv("../Lab3/AirQualityData(3).csv", header=TRUE, as.is=TRUE, sep=';')

# In this data set missing values are stored as -200
# Replace missing data (-200) with NA
aq_data[aq_data == -200] <- NA
# Note that this replaces -200 to NA across all columns in aq_data

# Create a new data frame 'tempdataset' by selecting specific columns from 'aq_data'
# For 'Time' column: extract the numeric components from a ...
# ... character string in a specific format and convert it into a numeric value. 
tempdataset <- data.frame(Time = as.numeric(sub("^(\\d+).(\\d+).*", "\\1.\\2",
                                                aq_data$Time)),
                          Temperature = aq_data$Temperature,
                          AbsoluteHumidity = aq_data$AbsoluteHumidity,
                          RelativeHumidity = aq_data$RelativeHumidity,
                          SensorCO = aq_data$PT08.S1.CO.,
                          GroundCO = aq_data$CO.GT.)

# Remove rows with missing values (NA)
Dataset <- tempdataset[complete.cases(tempdataset), ]

# Remove the 'tempdataset' data frame from the R environment
rm(tempdataset)
```

# Model Building

## Model Specifications

Ordinary least squares model

```{r ols-model, include=TRUE}
# Ordinary LS
m.ols <- lm(GroundCO ~ SensorCO, data = Dataset)
```

Quadratic least squares model

```{r quad-model, include=TRUE}
# Quadratic LS
m.quadls <- lm(GroundCO ~ SensorCO + I(SensorCO^2), data = Dataset)
```

Quartic  least squares model

```{r quartic-model, include=TRUE}
# Quartic LS
m.quartls <- lm(GroundCO ~ SensorCO + I(SensorCO^2) + 
                  I(SensorCO^3) + I(SensorCO^4), data = Dataset)
```

## Model Evaluation

Fitted values

```{r predictions, include=TRUE}
# Predictions/fitted-values for all three models
SensorCONew <- seq(600, 2200, len = length(Dataset$SensorCO))
predLinear <- predict(m.ols, newdata = data.frame(SensorCO = SensorCONew))
predQuad <- predict(m.quadls, newdata = data.frame(SensorCO = SensorCONew))
predQuart <- predict(m.quartls, newdata = data.frame(SensorCO = SensorCONew))
```

Fitted values plot

```{r fitted-values-plot, include=TRUE}
# Fitted values plot
# Add a scatterplot layer to the plot with specified point size (0.1)
# Add a line plot layer to the plot using the predLinear values, coloring them blue
# Add second line plot layer using the predQuad values coloring them red
# Add third  line plot layer using the predQuart values coloring them green
# Set the color scale for the lines with labels "Linear", "Quartic", ...
# ... and "Quadratic" and name it "Prediction" in the legend
# Set the title of the plot to "LS Predictions"
ggplot(Dataset, aes(x = SensorCO, y = GroundCO)) + 
  geom_point(size = 0.1) +
  geom_line(mapping = aes(x = SensorCONew, y = predLinear, color = "blue")) +
  geom_line(mapping = aes(x = SensorCONew, y = predQuad, color = "red")) +
  geom_line(mapping = aes(x = SensorCONew, y = predQuart, color = "green")) +
  scale_color_discrete(name = "Prediction", labels = c("Linear", "Quartic", "Quadratic")) +
  ggtitle("LS Predictions")
```

Standardized residuals for OLS

```{r residuals-ols, include=TRUE}
# Standardized Residual, Ordinary LS Standard Residual vs SensorCO
# Calculate the standardized residuals for the OLS model
StanResLS <- rstandard(m.ols)
# Create a new data frame containing two columns: SensorCO and StanResLS
dataLS <- data.frame(SensorCO = Dataset$SensorCO, StanResLS)
# Create a scatterplot using the data in dataLS, with SensorCO ...
# ... on the x-axis and StanResLS on the y-axis
# Set the point size to 0.1 for better visibility
# Add horizontal lines at y = 2 and y = -2 in blue to indicate ...
# ... the threshold for standardized residuals
# Set the title of the plot to "Standardized Residuals for OLS"
ggplot(dataLS, aes(x = SensorCO, y = StanResLS)) + 
  geom_point(size = 0.1) +
  geom_hline(yintercept = 2, color = 'blue') + 
  geom_hline(yintercept = -2, color = 'blue') +
  ggtitle("Standardized Residuals for OLS")
```

Standardized residuals for Quadratic LS

```{r residuals-quad, include=TRUE}
# Quadratic LS Standard Residual vs SensorCO
# Calculate standardized residuals for the quadratic linear model (m.quadls)
StanResQLS <- rstandard(m.quadls)
# Create a data frame 'dataQLS' with two columns: 'SensorCO' and 'StanResQLS'
dataQLS <- data.frame(SensorCO = Dataset$SensorCO, StanResQLS)
# Create a ggplot scatterplot with 'SensorCO' on the x-axis and 'StanResQLS' on the y-axis
# Add points to the plot with a specified point size (0.1)
# Add horizontal lines at y = 2 and y = -2 in blue to indicate threshold lines
# Set the title of the plot to "Standardized Residuals"
ggplot(dataQLS, aes(x = SensorCO, y = StanResQLS)) + 
  geom_point(size = 0.1) +
  geom_hline(yintercept = 2, color = 'blue') + 
  geom_hline(yintercept = -2, color = 'blue') +
  ggtitle("Standardized Residuals")
```

Combined standardized residuals plot

```{r combined-residuals-plot, include=TRUE}
# Combine LS and Quadratic Stan. Res. plot in the same
# Create a ggplot scatterplot with 'SensorCO' on the x-axis and 'StanResQLS' ...
# ... as well as 'StanResQLS' on the y-axis
# Add points to the plot with a specified point size (0.1)
# Add horizontal lines at y = 2 and y = -2 in blue to indicate threshold lines
# Set a manual color scale for the points with labels "Linear" and "Quadratic"
# Add a label for the y-axis
# Set the title of the plot to "Standardized Residuals"
ggplot() + 
  geom_point(data = dataLS, aes(x = SensorCO, y = StanResLS, color = "Linear"),
             size = 0.1) +
  geom_point(data = dataQLS, aes(x = SensorCO, y = StanResQLS, color = "Quadratic"),
             size = 0.1) +
  geom_hline(yintercept = 2, color = 'blue') + 
  geom_hline(yintercept = -2, color = 'blue') +
  scale_color_manual(name = element_blank(), labels = c("Linear", "Quadratic"),
                     values = c("blue", "red")) +
  labs(y = "Standardized Residual") + 
  ggtitle("Standardized Residuals Plot") 
```

Standardized residuals vs fitted for OLS and Quadratic LS

```{r residuals-vs-fitted, include=TRUE}
# Standardized Residuals vs Fitted
# Calculate the fitted values for the OLS model
Fitted <- fitted(m.ols)
# Create a data frame containing the fitted values and the standardized residuals
dataLSFitted <- data.frame(Fitted, StanResLS)
# Calculate the fitted values for the QuadLS model
Fitted <- fitted(m.quadls)
# Create a data frame containing the fitted values and the standardized residuals
dataQLSFitted <- data.frame(Fitted, StanResQLS)

# Combine OLS and Quadratic LS standardized residuals vs fitted values plot 
# Create a ggplot scatterplot with fitted values on the x-axis and 'StanResQLS' ...
# ... as well as 'StanResQLS' on the y-axis
# Add points to the plot with a specified point size (0.1)
# Add horizontal lines at y = 2 and y = -2 in blue to indicate threshold lines
# Set a manual color scale for the points with labels "Linear" and "Quadratic"
# Add a label for the y-axis
# Set the title of the plot to "Standardized Residuals"
ggplot() + 
  geom_point(data = dataLSFitted, aes(x = Fitted, y = StanResLS, color = "Linear"),
             size = 0.1) +
  geom_point(data = dataQLSFitted, aes(x = Fitted, y = StanResQLS, color = "Quadratic"),
             size = 0.1) +
  geom_hline(yintercept = 2, color = 'blue') + 
  geom_hline(yintercept = -2, color = 'blue') +
  scale_color_manual(name = element_blank(), labels = c("Linear", "Quadratic"),
                     values = c("blue", "red")) +
  labs(y = "Standardized Residual") + 
  labs(x = "Fitted value") + 
  ggtitle("Standardized Residuals Plot (Fitted) ") 
```

Scatterplot matrix with correlation values

```{r scatterplot-matrix, include=TRUE}
# Scatterplot Matrix with Correlation Values
# Create a scatterplot matrix (pairs plot) for the 'Dataset' data frame
# The 'upper' panel contains correlation values with a point size of 4
# This visualization helps to understand variable relationships and correlations
# The 'lower' panel contains scatterplots with points and alpha transparency of 0.3
# The point size in the upper panel is set to 0.1
ggpairs(Dataset,
        upper = list(continuous = wrap('cor', size = 4)),
        lower = list(continuous = wrap("points", alpha = 0.3, size = 0.1)))
```

# Multiple Linear Regression (MLR)

Building a multiple linear regression model

```{r mlr-model, include=TRUE}
# Create a Multiple Linear Regression Model
# Model 'GroundCO' variable using multiple predictors including ...
# ... 'SensorCO', 'Temperature', 'RelativeHumidity', 'AbsoluteHumidity', ...
# ... and 'Time' from the 'Dataset' data frame.
m.mlr <- lm(GroundCO ~ SensorCO + Temperature + RelativeHumidity +
              AbsoluteHumidity + Time,
            data = Dataset)
```

Summary of the MLR model

```{r mlr-summary, include=TRUE}
# Generate a summary of the Multiple Linear Regression (MLR) model
summary(m.mlr)
```

Standard residuals for MLR

```{r residuals-mlr, include=TRUE}
# Standard Residuals vs SensorCO
# Examine Standardized residuals
# 'rstandard' calculates the standardized residuals for a linear regression model
StanResMLR <- rstandard(m.mlr)
# Create a data frame containing the SensorCO values and the standardized residuals
dataMLS <- data.frame(SensorCO = Dataset$SensorCO, StanResMLR)

# Combine Quad LS and MLR standardized residuals vs SensorCO plot 
# Create a ggplot scatterplot with SensorCO on the x-axis and 'StanResQLS' ...
# ... as well as 'StanResMLR' on the y-axis
# Add points to the plot with a specified point size (0.1)
# Add horizontal lines at y = 2 and y = -2 in blue to indicate threshold lines
# Set a manual color scale for the points with labels "Linear" and "Quadratic"
# Add a label for the y-axis
# Set the title of the plot to "Standardized Residuals"
ggplot() + 
  geom_point(data = dataQLS, aes(x = SensorCO, y = StanResQLS, color = "Quadratic"),
             size = 0.1) +
  geom_point(data = dataMLS, aes(x = SensorCO, y = StanResMLR, color = "MLR"),
             size = 0.1) +
  geom_hline(yintercept = 2, color = 'blue') + 
  geom_hline(yintercept = -2, color = 'blue') +
  scale_color_manual(name = element_blank(), labels = c("MLR", "Quadratic"),
                     values = c("red", "blue")) +
  labs(y = "Standardized Residual") + 
  ggtitle("Standardized Residuals Plot")
```

Standardized residuals vs fitted for MLR and Quadratic LS

```{r residuals-vs-fitted-mlr, include=TRUE}
# Standardized Residuals vs Fitted
# Calculate the fitted values for the Quad LS model
Fitted <- fitted(m.quadls)
# Create a data frame containing the fitted values and the standardized residuals
dataQLSFitted <- data.frame(Fitted, StanResQLS)
# Calculate the fitted values for the MLR model
Fitted <- fitted(m.mlr)
# Create a data frame containing the fitted values and the standardized residuals
dataMLRFitted <- data.frame(Fitted, StanResMLR)

# Combine Quadratic and MLR standardized residuals vs fitted values plot 
# Create a ggplot scatterplot with fitted values on the x-axis and 'StanResQLS' ...
# ... as well as 'StanResMLR' on the y-axis
# Add points to the plot with a specified point size (0.1)
# Add horizontal lines at y = 2 and y = -2 in blue to indicate threshold lines
# Set a manual color scale for the points with labels "MLR" and "Quadratic"
# Add a label for the y-axis
# Set the title of the plot to "Standardized Residuals"
ggplot() + 
  geom_point(data = dataQLSFitted, aes(x = Fitted, y = StanResQLS, color = "Quadratic"),
             size = 0.1) +
  geom_point(data = dataMLRFitted, aes(x = Fitted, y = StanResMLR, color = "MLR"),
             size = 0.1) +
  geom_hline(yintercept = 2, color = 'blue') + 
  geom_hline(yintercept = -2, color = 'blue') +
  scale_color_manual(name = element_blank(), labels = c("MLR", "Quadratic"),
                     values = c("red", "blue")) +
  labs(y = "Standardized Residual") + 
  labs(x = "Fitted value") + 
  ggtitle("Standardized Residuals Plot (Fitted) ") 
```

## Comparing Models

Comparison with Quadratic LS

Building a quadratic MLR model

```{r quad-mlr-model, include=TRUE}
# Create a Multiple Linear Regression Model
# Model 'GroundCO' variable using multiple predictors including ...
# ... 'SensorCO', 'SensorCO^2', 'Temperature', 'RelativeHumidity', 'AbsoluteHumidity', ...
# ... and 'Time' from the 'Dataset' data frame.
mquad.mls <- lm(GroundCO ~ SensorCO + I(SensorCO^2) + Temperature + RelativeHumidity +
                  AbsoluteHumidity + Time,
                data = Dataset)
```

Summary of the quadratic MLR model

```{r quad-mlr-summary, include=TRUE}
# Generate a summary of the mquad.mls model
summary(mquad.mls)
```

Compare with quadratic model
```{r compare-models, include=TRUE}
# Compare with the summary of m.quadls model
summary(m.quadls)
```